Studies of quantum dots

### 2016-12-14 Manuscript/paper.tex
Add some results and figures to the manuscript

\input{head_plain}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{color}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{siunitx}
\usepackage{subfigure}

\newcommand{\D}{\operatorname{d\!}}
\newcommand{\bra}[1]{\langle #1 |}
\newcommand{\braket}[2]{\langle #1 | #2\rangle}
\newcommand{\ket}[1]{| #1 \rangle}
\newcommand{\normord}[1]{\mathopen: #1 \mathclose:}

@@

An accurate and robust ODE solver is required to solve \eqref{eq:imsrgode}.  In particular, the solver must be able to handle the stiffness that can often arise (even in the case of White generator).  For this, we use an accurate, high-order ODE solver implementation by Shampine and Gordon that uses the Adams predictor-corrector formulas [[Ref: Shampine and Gordon book]].

@@

The commutator in the flow equations \eqref{eq:imsrgode} guarantees that the evolved state $\hat U(s) \ket{\Phi}$ can be expanded in terms of \emph{linked diagrams} only \cite{shavitt2009many,ISI:A1981MN73700014}, which demonstrates that IM-SRG is size-extensive.  Regarding the quality of the SRG results, it means that the error introduced by truncating the many-body expansions scales linearly with the number of particles $N$.

@@

Particle attached and particle removed equations of motion (EOM) methods can be coupled with either IM-SRG or coupled cluster calculations. The principal idea is that one may construct a ladder operator $\hat{X}$ that promotes the $N$-particle ground state to any state in the $N + 1$ or $N - 1$ spectrum,
\begin{equation}\label{eq:ladder_def}
  \ket{\Psi^{(N \pm 1)}_u}  = \hat{X}^{(N \pm 1)}_u \ket{\Psi^{(N)}_0},
\end{equation}
where $\hat{X}$ is in principle a linear combination of excitation and de-excitation operators that change particle number by one,
\begin{align}
  \label{eq:gen_attached}
  \hat{X}^{(N+1)}_u &= \sum_p \bar{n}_p x^{[1,0]}_p  \hat{a}^\dagger_p + \frac{1}{2} \sum^{[2,1]}_{p, q, r} \bar{n}_p \bar{n}_q n_r x_{p, q, r} \hat{a}^\dagger_p \hat{a}^\dagger_q \hat{a}_r + \cdots,  \\
  \label{eq:gen_removed}
  \hat{X}^{(N-1)}_u &= \sum_p n_p x_p^{[0,1]} \hat{a}_p + \frac{1}{2} \sum_{p, q, r} n_p n_q \bar{n}_r  x^{[1,2]}_{p, q, r} \hat{a}^\dagger_r \hat{a}_q \hat{a}_p  + \cdots.
\end{align}
Substitution of \eqref{eq:ladder_def} into the energy eigenvalue problem $\hat H \ket{\Psi_u^{(N \pm 1)}} = E^{(N\pm1)}_u \ket{\Psi_u^{(N \pm 1)}}$ gives
\begin{equation}\label{eq:EOM}
  [\hat{H}, \hat{X}^{(N \pm 1)}_u] \ket{\Psi^{(N)}_0} = \pm \epsilon^{(\pm)}_u \hat{X}^{(N \pm 1)}_u \ket{\Psi^{(N)}_0},
\end{equation}
which constitutes a generalized eigenvalue problem for the amplitudes $x$, where $\epsilon^{(\pm)}_u$ are the single-particle addition ($+$) and removal ($-$) energies. The quality of this calculation depends on the ansatz for the $N$-particle ground state, as well as the systematically improvable truncation on the ladder operators. In this work we include 1p and 2p1h excitations in the $N + 1$ ladder operator and likewise 1h and 2h1p operators for the $N - 1$ ladder operators.

@@

[[begin unmerged]]

In this section, we discuss the IM-SRG results for the ground-state energies of closed-shell systems. In subsection \ref{subsec:Wegner}, we start with the Wegner generator. Numerical instabilities in the integration process, as well as a rather slow convergence as function of the number of oscillator shells, motivate the use of a Hartree-Fock basis and the Coulomb interaction. Since higher correlations still lead to stiff equation systems, we change in subsection \ref{subsec:White} to the White generator. In particular, we give an overview of the IM-SRG(2) ground state energies for closed-shell systems up to $N=42$ particles and compare with other \textit{ab initio} many-body methods.

In all our calculations, we perform the integration of the flow equations using the algorithm developed by Shampine and Gordon \cite{shampine1975computer}. This algorithm solves linear systems of differential equations based on the implicit Adams method, which is a multi-step method of variable order. Source code is freely available \cite{odesolver}.

[[end unmerged]]

In Figure [[fig:gs]] we show the calculated ground state energies from
Hartree-Fock, second-order (M\o ller-Plesset) perturbation theory (MPPT), and
IM-SRG.  The exact result from diffusion Monte Carlo (DMC) is also shown as a
dashed line in systems where it is known \cite{PhysRevB.84.115302}.

@@

There are a few cases where the IM-SRG over-corrects the result, leading to a
ground state energy that is lower than the exact result.  This is not
unexpected given that, unlike Hartree-Fock, IM-SRG is non-variational.  These
tend to occur when the number of particles is few, or the frequency is very
low (high correlation).  However, MPPT does not do much better in the high
frequency regimes as it often converges slower than IM-SRG.

[[Q: should we discuss the absolute/relative correction plots?]]

For our addition and removal energy calculations, the results are summarized
in Figure [[fig:add]] and Figure [[fig:rm]] respectively.  The figures
show the the addition/removal energies for Hartree-Fock, combined with IM-SRG
and/or QDPT (Hartree-Fock is always included, as usual) [[better way to
explain this? the figure labels are somewhat misleading too]].  The DMC
results are also shown via dashed lines where they are known.

@@

For the IM-SRG results, there is a small asymmetry between the behavior of our
removal and addition energies: we find that removal energies tend to be more
accurate and its behavior is relatively stable with respect to the number of
shells. [[why]]

@@

We also note that IM-SRG with 2 particles and frequency 0.1 generally fail to
give any result as the ODE solver eventually leads to divergence [[Why?]]
This seems to be a particular quirk associated with this specific system and
does not usually occur in other systems except in a few extreme, isolated
cases where there are far too few unoccupied states.

@@

We note that this calculation was done entirely using the traditional approach of using a high-order ODE solver to solve the flow equation.  A new technique developed by T. Morris \cite{PhysRevC.92.034331} uses an alternative approach that obviates the need for a high-order ODE solver, leading to much more efficient computations and also allowing operators of other observables to be evolved with greater ease.  Implementing this approach would be beneficial as it would allows us to study the accuracy and convergence of other observables beyond energy and energy differences.

>>>>

\input{head_plain.tex}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{color}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{siunitx}
\usepackage{subfigure}

\newcommand{\D}{\operatorname{d\!}}
\newcommand{\E}{\operatorname{e}}
\newcommand{\bra}[1]{\langle #1 |}
\newcommand{\braket}[2]{\langle #1 | #2\rangle}
\newcommand{\ket}[1]{| #1 \rangle}
\newcommand{\normord}[1]{\mathopen: #1 \mathclose:}

@@

An accurate and robust ODE solver is required to solve \eqref{eq:imsrgode}.  In particular, the solver must be able to handle the stiffness that can often arise (even in the case of White generator, which is designed to avoid some of the stiffness).  For our numerical experiments, we used a high-order ODE solver algorithm by L.\ F.\ Shampine and M.\ K.\ Gordon \cite{shampine1975computer}, which is a multistep method based on the implicit Adams predictor-corrector formulas.  Its source code is freely available \cite{odesolver}.

@@

The commutator in the flow equations \eqref{eq:imsrgode} indicates that the evolved state $\hat U(s) \ket{\Phi}$ can be expanded in terms of \emph{linked diagrams} only \cite{shavitt2009many,ISI:A1981MN73700014}, which ensures the size-extensive property of IM-SRG by construction.  This holds even if the commutator is truncated.  Regarding the quality of the SRG results, this indicates that the error introduced by truncating the many-body expansions scales linearly with the number of particles $N$.

@@

Particle attached and particle removed equations of motion (EOM) methods can be coupled with either IM-SRG or coupled cluster calculations. The principal idea is that one may construct a ladder operator $\hat{X}$ that promotes the $N$-particle ground state to any state in the $N + 1$ or $N - 1$ spectrum,
\begin{equation}\label{eq:ladder_def}
  \ket{\Psi^{(N \pm 1)}_u}  = \hat{X}^{(N \pm 1)}_u \ket{\Psi^{(N)}_0},
\end{equation}
where $\hat{X}$ is in principle a linear combination of excitation and de-excitation operators that change particle number by one,
\begin{align}
  \label{eq:gen_attached}
  \hat{X}^{(N+1)}_u &= \sum_p \bar{n}_p x^{[1,0]}_p  \hat{a}^\dagger_p + \frac{1}{2} \sum^{[2,1]}_{p, q, r} \bar{n}_p \bar{n}_q n_r x_{p, q, r} \hat{a}^\dagger_p \hat{a}^\dagger_q \hat{a}_r + \cdots,  \\
  \label{eq:gen_removed}
  \hat{X}^{(N-1)}_u &= \sum_p n_p x_p^{[0,1]} \hat{a}_p + \frac{1}{2} \sum_{p, q, r} n_p n_q \bar{n}_r  x^{[1,2]}_{p, q, r} \hat{a}^\dagger_r \hat{a}_q \hat{a}_p  + \cdots.
\end{align}
Substitution of \eqref{eq:ladder_def} into the energy eigenvalue problem
\begin{gather*}
  \hat H \ket{\Psi_u^{(N \pm 1)}} = E^{(N\pm1)}_u \ket{\Psi_u^{(N \pm 1)}}
\end{gather*}
gives
\begin{equation}\label{eq:EOM}
  [\hat{H}, \hat{X}^{(N \pm 1)}_u] \ket{\Psi^{(N)}_0} = \pm \epsilon^{(\pm)}_u \hat{X}^{(N \pm 1)}_u \ket{\Psi^{(N)}_0},
\end{equation}
which constitutes a generalized eigenvalue problem for the amplitudes $x$, where $\epsilon^{(\pm)}_u$ are the single-particle addition ($+$) and removal ($-$) energies. The quality of this calculation depends on the ansatz for the $N$-particle ground state, as well as the systematically improvable truncation on the ladder operators. In this work we include 1p and 2p1h excitations in the $N + 1$ ladder operator and likewise 1h and 2h1p operators for the $N - 1$ ladder operators.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{fig-gs}
  \caption{Ground state energies.}
  \label{fig:gs}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{fig-add}
  \caption{Addition energies.}
  \label{fig:add}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{fig-rm}
  \caption{Removal energies.}
  \label{fig:rm}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{fig-rel-slopes}
  \caption{Comparison of the rate of convergence for various modifications to the Coulomb interaction.  The key observation from this plot is that the short-range part of the interaction has a strong impact on the convergence, whereas the long-range tail has little effect.}
  \label{fig:rel-slopes}
\end{figure}

In this section, we discuss the IM-SRG results for the ground-state energies of closed-shell systems . In subsection \ref{subsec:Wegner}, we start with the Wegner generator. Numerical instabilities in the integration process, as well as a rather slow convergence as function of the number of oscillator shells, motivate the use of a Hartree-Fock basis and the Coulomb interaction.  Since higher correlations still lead to stiff equation systems, we change in subsection \ref{subsec:White} to the White generator. In particular, we give an overview of the IM-SRG(2) ground state energies for closed-shell systems up to $N=42$ particles and compare with other \textit{ab initio} many-body methods.

[[...]]

In Figure \ref{fig:gs} we show the calculated ground state energies from Hartree-Fock, second-order (M\o ller-Plesset) perturbation theory (MPPT), and IM-SRG.  The exact result from diffusion Monte Carlo (DMC) is also shown as a dashed line in systems where it is known \cite{PhysRevB.84.115302}.

@@

There are a few cases where the IM-SRG over-corrects the result, leading to a ground state energy that is lower than the exact result.  This is not unexpected given that, unlike Hartree-Fock, IM-SRG is non-variational in the presence of operator truncations that cause small unitarity violations.  Such behavior tends to occur when the number of particles is few, or the frequency is very low (high correlation).  However, MPPT does not do much better in the high frequency regimes as it often converges slower than IM-SRG.

For our addition and removal energy calculations, the results are summarized in Figure \ref{fig:add} and Figure \ref{fig:rm} respectively.  The figures show the the addition/removal energies for Hartree-Fock, combined with IM-SRG and/or QDPT (Hartree-Fock is always included, as usual) [[better way to explain this? the figure labels are somewhat misleading too]].  The DMC results are also shown via dashed lines where they are known.

@@

@@

We also note that IM-SRG with 2 particles and a frequency 0.1 generally fail to give any result as the ODE solver eventually leads to divergence [[Why?]]  This seems to be a quirk isolated to this particular system and does not usually occur in other systems.

[[---]]

To investigate the drifts, we tested our methods against a modified form of the Coulomb interaction, parametrized by two lengths $\sigma_{\mathrm{A}}$ and $\sigma_{\mathrm{B}}$ that characterize the range of the interaction:
\begin{align}
  V_{\sigma_{\mathrm{A}}, \sigma_{\mathrm{B}}}(r) \equiv \frac{(1 + c)^{1 - 1/c}}{c} \left(1 - \E^{-r^2 / (2 \sigma_{\mathrm{A}}^2)}\right) \E^{-r^2 / (2 \sigma_{\mathrm{B}}^2)} \frac{1}{r}
\end{align}
where $c \equiv \sqrt{\sigma_{\mathrm{B}} / \sigma_{\mathrm{A}}}$.  The coefficient is defined to ensure the peak of the envelope remains at unity.  By setting $\sigma_{\mathrm{A}} = 0$ and $\sigma_{\mathrm{B}} = \infty$ one recovers the original Coulomb interaction.  By increasing $\sigma_{\mathrm{A}}$ one can reduce the short-range part of the interaction, and analogously by increasing $\sigma_{\mathrm{B}}$ one can reduce the long-range part of the interaction.  For our numerical experiments we consider the following four combinations of $(\sigma_{\mathrm{A}}, \sigma_{\mathrm{B}})$: $(0, \infty)$, $(\frac{1}{2}, \infty)$, $(0, 4)$, $(\frac{1}{2}, 4)$.

To analyze the convergence as function of the number of shells $K$, we use a simple but approximate metric, the \textit{relative slope} at $K = 15$:
\begin{align*}
  \rho_{15}(\varepsilon) \equiv \frac{\varepsilon_{K = 15} - \varepsilon_{K = 14}}{\varepsilon_{K = 15}}
\end{align*}
For the systems we have selected, at $K = 15$ the addition and removal energies have essentially reached the asymptotic behavior (by visual inspection), thus $\rho_{15}$ gives a crude measure of the rate of convergence as a relative quantity independent of the energy unit, allowing it to be meaningfully compared between systems with different interactions and different particle numbers.  The results are plotted in Figure \ref{fig:rel-slopes}.

Analyzing the figure, we observe that the reducing short-range part of the interaction improves the rate of convergence greatly -- the relative slope has saturated and reached the precision of the ODE solver -- whereas eliminating the long-range part of the interaction had very little effect.  This demonstrates that the main cause of the slow convergence lies in the highly repulsive, short-ranged part of the interaction, which leads to the presence of nondifferentiable cusps (the so-called \textit{Coulomb cusps}) in the exact wave functions that are difficult to reproduce exactly using linear combinations of the smooth harmonic oscillator wave functions.

The figure also tells us that the convergence is negatively impacted at lower frequencies and, to a lesser extent, by the number of particles.  Both are expected: lower frequencies increase the correlation in the system, while higher number of particles naturally require more shells to converge.

We also note that the behavior of addition energies versus removal energies tend to be similar, and which one converges faster fluctuates from system to system.

@@

We note that this calculation was done entirely using the traditional approach of using a high-order ODE solver to solve the flow equation.  A new technique developed by T. D. Morris \cite{PhysRevC.92.034331} uses an alternative approach that obviates the need for a high-order ODE solver, leading to much more efficient computations and also allowing operators of other observables to be evolved with greater ease.  Implementing this approach would be beneficial as it would allows us to study the accuracy and convergence of other observables beyond energy and energy differences.
